Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-04-21 draft_BLC/BLC_AM-ID_manuscript.tex
revision minor edits

Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent. In India, it is spoken as an L2 by 129 million people (Census of India, 2011).  English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
%citations 
Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review). This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English. Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target. Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.

%%%%%%%%%%%%%%%%%%%%%%%do something%%%%%%%%%%%%%%%
Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:

@@

@@

Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time. The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}. However, longitudinal studies %(citations) 
and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts. For example, \cite[][vowel quality]{bohn1992production} 
observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer. These have been interpreted as changes to long-term phonological representations.

@@

@@

Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status. However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience. The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories. As far as we are aware, there is no work yet on vowels in L2. Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.

@@

Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions. Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure. In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too. Once again, this emphasizes the importance of examining a wider range of sound contrasts.

@@

In the present study, we do not elicit unilingual Bengali productions. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used Bengali production data from five female speakers from \cite{dutta2021}. Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel. Next, we compared these to our participants' productions in the unilingual English condition. Comparisons of Lobanov-normalized F1 and F2 suggest that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.

@@

Power is a function of sample size and effect size. Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}. This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$. Effect sizes in simr are expressed as unstandardized effect sizes (model estimates). In the absence (to the best of our knowledge) of literature on typical effect sizes for spectral differences in cross-language vowel production, we made a conservative estimate of effect sizes as |$\beta$| > 0.1. This amounts to the assumption that the smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}) is 0.1, and we want to confirm whether the study has enough power to detect such an effect, if it were to exist. Note that this is an arbitrary assumption; only more discussion among researchers in this field will lead to a consensus on what is considered a typical cross-language effect. 
In the  Calculations of d values from our models using the package EMAtools \citep{ematools} show that they range from 0.01 to 1.06 (no need).

>>>>

Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent. In India, it is spoken as an L2 by 129 million people (Census of India, 2011).  English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.

\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested. For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}

Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review). This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English. Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target. Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.

Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:

@@

\alert{discussion: this confirms that the local act of switching does cause transfer, beyond language mode. Most existing studies looking at local effect of switching measure VOT. We find that the sae happens in vowels. Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift. Because vowel quality measures have directionality as well as quantity, we can see finer patterns. We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category. What does this say about what happens during online processing and speech planning? --- look at discussion section of the mode vs switching papers: what do they say about this? Can this finding be accounted for with that?}

\alert{It makes sense that a bilingual language mode makes the representations from boh languages available. But how are these processed during the act of using language? If these representtions are combined in creative ways, then are we moving between languages during a single discourse? Can we think of intermediate representations, not of individual sounds, but rather of sound systems? }

@@

Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time. The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}. However, longitudinal studies %(citations) 
and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts. For example, \cite[][vowel quality]{bohn1992production} 
observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence. These have been interpreted as changes to long-term phonological representations.

@@

\alert{discussion: found that like VOT, vowel quality shifts due to switching are also highly localized, and reset in an order of seconds. This behavior is similar to other context-driven shifts in the literature}

@@

Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status. However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience. The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories. As far as we are aware, there is no work yet on short-term transfer in L2 vowels. Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.

\alert{discussion: more about ICM, assumptions about processing}

@@

Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions. Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure. In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' and contrast constrains transfer, then we should expect it to apply to individual sounds too. Once again, this emphasizes the importance of examining a wider range of sound contrasts.

@@

\alert{in discussion- say: fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shif in the common category because it's "confusing". But this-- shows that the L1 influence is at a system-level. More inclined towards using a Bengali-like phonological system. Has implications for cognitive models of language and what happens during short-term interaction)}

In the present study, we do not elicit unilingual Bengali productions. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used productions from a freely available Bengali corpus SHRUTI \citep{shruticorpus}. Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel. Next, we compared these to our participants' productions in the unilingual English condition. \alert{fig. .. plots} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.

@@

Power is a function of sample size and effect size. Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}. This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$. Effect sizes in simr are expressed as unstandardized effect sizes (model estimates). In the absence (to the best of our knowledge) of literature on typical effect sizes for spectral differences in cross-language vowel production, we made a conservative estimate of effect sizes as |$\beta$| > 0.1. This amounts to the assumption that the smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}) is 0.1, and we want to confirm whether the study has enough power to detect such an effect, if it were to exist. Note that this is an arbitrary assumption; only more discussion among researchers in this field will lead to a consensus on what is considered a typical cross-language effect.

"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients. relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations." \citep{kumle2021estimating}

Next, following \cite{kumle2021estimating}, we ran a power analysis assuming smaller effect sizes by reducing the beta-coefficients by 15\%. (Use the model summary, make a vector of all beta-values, including intercept, add as argument of SESOI= )

In the  Calculations of d values from our models using the package EMAtools \citep{ematools} show that they range from 0.01 to 1.06 (no need).
